model_fit

### 2025-12-07 model_fit.Rmd
added models that got deleted and make sure code worked, and graphics show

```{r, results='hide'}
library(readr)
library(tidyverse)
library(boot)
library(glmnet)
library(caret)
library(pROC)
library(MASS)
library(ggplot2)

@@

## 3.4 K-Nearest Neighbors (KNN)
- Standardize numeric predictors.
- Perform CV to tune **k** .
- Report best k and corresponding CV error.
- Discuss bias-variance behavior.

## 3.5 Generalized Additive Models (GAM)
- Fit GAM with smoothing splines for continuous variables.
- Use CV to choose smoothing parameters.
- Examine partial effect plots for interpretability.

## 3.6 Boosting (Most Flexible Model)
- Fit gradient boosting classifier (e.g., `gbm` or `xgboost`).
- Use cross-validation for:
  - number of trees,
  - interaction depth,
  - learning rate.
- Discuss complexity vs interpretability tradeoff. This model will be our benchmark non-interpretable model to compare interpretable models fitted above.

@@

@@

>>>>

```{r, results='hide'}
library(readr)
library(tidyverse)
library(boot)
library(glmnet)
library(caret)
library(pROC)
library(MASS)
library(ggplot2)
library(dplyr)

@@

- Standardize numeric predictors.
- Perform CV to tune **k** .
- Report best k and corresponding CV error.
- Discuss bias-variance behavior.

## 3.4 K-Nearest Neighbors (KNN)
KNN is a nonparametric method that relies on feature scaling and distance-based classification.

Prepare scaled numeric predictors
```{r}
library(class)

numeric_vars <- names(train_data)[sapply(train_data, is.numeric)]
numeric_vars <- setdiff(numeric_vars, "residuals")

train_x <- train_data[, numeric_vars]
test_x  <- test_data[, numeric_vars]

train_x_scaled <- scale(train_x)
test_x_scaled <- scale(
  test_x,
  center = attr(train_x_scaled, "scaled:center"),
  scale  = attr(train_x_scaled, "scaled:scale")
)

train_y <- train_data$y
test_y  <- test_data$y

Cross-validation to choose the best k

```{r}
k_values <- seq(1, 25, 2)
cv_accuracy <- numeric(length(k_values))

set.seed(445)
folds <- createFolds(train_y, k = 5)

for (i in seq_along(k_values)) {
  k <- k_values[i]
  fold_acc <- c()

  for (f in folds) {
    pred <- knn(
      train = train_x_scaled[-f, ],
      test  = train_x_scaled[f, ],
      cl    = train_y[-f],
      k = k
    )
    fold_acc <- c(fold_acc, mean(pred == train_y[f]))
  }

  cv_accuracy[i] <- mean(fold_acc)
}

best_k <- k_values[which.max(cv_accuracy)]
best_k

Final KNN prediction and metrics

```{r}
knn_pred <- knn(
  train = train_x_scaled,
  test  = test_x_scaled,
  cl    = train_y,
  k = best_k
)

cm_knn <- confusionMatrix(knn_pred, test_y, positive = "yes")
cm_knn

knn_prob <- as.numeric(knn_pred == "yes")
roc_knn <- roc(test_y, knn_prob)
auc_knn <- auc(roc_knn)
auc_knn

## 3.5 Generalized Additive Models (GAM)

GAM allows flexible nonlinear fits through smoothing splines.

```{r}
library(gam)

gam_model <- gam(
  y ~ s(age) + s(balance) + s(duration) + s(campaign) + s(previous),
  data = train_data,
  family = binomial
)

summary(gam_model)

gam_prob <- predict(gam_model, test_data, type = "response")
gam_pred <- ifelse(gam_prob > 0.5, "yes", "no") |> factor(levels = c("no","yes"))

cm_gam <- confusionMatrix(gam_pred, test_y, positive = "yes")
cm_gam

roc_gam <- roc(test_y, gam_prob)
auc_gam <- auc(roc_gam)
auc_gam

## 3.6 Boosting (Most Flexible Model)

Boosting builds many small trees sequentially, improving mistakes made by prior trees.

```{r}

if ("residuals" %in% colnames(train_data)) {
  train_data$residuals <- NULL
}

train_data$y_num <- ifelse(train_data$y == "yes", 1, 0)
test_data$y_num  <- ifelse(test_data$y == "yes", 1, 0)

library(gbm)
set.seed(445)

boost_model <- gbm(
  y_num ~ . - y,         
  data = train_data,
  distribution = "bernoulli",
  n.trees = 400,       
  shrinkage = 0.01,
  interaction.depth = 3,
  bag.fraction = 0.5,
  verbose = FALSE
)

best_trees <- gbm.perf(boost_model, method = "OOB", plot.it = FALSE)
best_trees

Prediction + Metrics
```{r}
boost_prob <- predict(boost_model, test_data, n.trees = best_trees, type = "response")
boost_pred <- factor(ifelse(boost_prob > 0.5, "yes", "no"), levels = c("no","yes"))

cm_boost <- confusionMatrix(boost_pred, test_y, positive = "yes")
cm_boost

roc_boost <- roc(test_y, boost_prob)
auc_boost <- auc(roc_boost)
auc_boost

@@

```{r}
final_results <- data.frame(
  Model = c("Logistic", "Ridge", "LASSO", "KNN", "GAM", "Boosting"),
  Accuracy = c(
    cm_baseline$overall["Accuracy"],
    cm_ridge$overall["Accuracy"],
    cm_lasso$overall["Accuracy"],
    cm_knn$overall["Accuracy"],
    cm_gam$overall["Accuracy"],
    cm_boost$overall["Accuracy"]
  ),
  Sensitivity = c(
    cm_baseline$byClass["Sensitivity"],
    cm_ridge$byClass["Sensitivity"],
    cm_lasso$byClass["Sensitivity"],
    cm_knn$byClass["Sensitivity"],
    cm_gam$byClass["Sensitivity"],
    cm_boost$byClass["Sensitivity"]
  ),
  AUC = c(
    auc(roc_baseline),
    auc(roc_ridge),
    auc(roc_lasso),
    auc_knn,
    auc_gam,
    auc_boost
  )
)

final_results

@@

Combined ROC Curve
```{r}
roc_all <- rbind(
  data.frame(fpr = 1 - roc_baseline$specificities, tpr = roc_baseline$sensitivities, Model = "Logistic"),
  data.frame(fpr = 1 - roc_ridge$specificities,    tpr = roc_ridge$sensitivities,    Model = "Ridge"),
  data.frame(fpr = 1 - roc_lasso$specificities,    tpr = roc_lasso$sensitivities,    Model = "LASSO"),
  data.frame(fpr = 1 - roc_knn$specificities,      tpr = roc_knn$sensitivities,      Model = "KNN"),
  data.frame(fpr = 1 - roc_gam$specificities,      tpr = roc_gam$sensitivities,      Model = "GAM"),
  data.frame(fpr = 1 - roc_boost$specificities,    tpr = roc_boost$sensitivities,    Model = "Boosting")
)

ggplot(roc_all, aes(x = fpr, y = tpr, color = Model)) +
  geom_line(size = 1) +
  geom_abline(linetype = "dashed") +
  labs(
    title = "ROC Curve Comparison for All Models",
    x = "False Positive Rate",
    y = "True Positive Rate"
  ) +
  theme_minimal()
